% <wc:start description="Safety v. Liability" max=800>

Since my undergraduate years a decade ago Harvard has struggled to balance
the pleasures and dangers of alcohol on campus. Many of the changes I've
witnessed are definite improvements. For me, AlcoholEdu was the evening I
spent crawling back and forth to the toilet, and my DAPA was my freshman
girlfriend cradling my head above it. I hadn't a clue about alcohol in high
school and arrived at college armed with little more than trial and error to
figure it out. A trip to the Ad Board (trial) and many mixed drinks (error)
later, I started to get the hang of things. Today we have an Office of
Alcohol and Other Drug Services which does a number of good things: providing
AlcoholEdu, running the DAPA program, employing Ryan Travia and offering
something called e-TOKE which I have not experimented with but have been
assured is not a joke.

Yet, despite these new programs, alcohol on campus remains dangerous, as
evidenced by the sobering statistics recently released about alcohol-related
hospital admissions this last fall. The College's refusal to release these
statistics along with information that might help put the numbers into
context is frustrating, and despite DAPA claiming to avoid scare tactics,
they administration did seem to feel the statistics were useful to scare the
Resident Tutors. At an unusual mid-year residential life staff meeting,
Provost Hyman and Dean Hammonds presented these statistics to Tutors from
across the college. Both left before the event concluded with a discussion
led by a panel of veteran tutors. After an Adams House tutor delivered a
stunning declaration of love for the party registration form, the panel came
to a general agreement that alcohol remains a problem and we should think
about how to address it.

% 11 Feb 2011 : GWA : From Seth:
%               By taking an unrealistic and paternalistic approach to
%               drinking on campus, the administration has driven students to
%               drink in more irresponsible ways. When I was an
%               undergraduate, we had many more opportunities to drink a few
%               beers at local bars or Stein Clubs whereas now students feel
%               they have to pound shots in their rooms. Every other
%               university in the nation I've ever been to takes a much more
%               reasoned and less confrontational approach to drinking,
%               whereas Harvard seems more concerned about being sued than
%               ensuring the growth and welfare of its students. Maybe these
%               other schools all just have better lawyers, but I suspect the
%               issue is that we, as campus leaders and administrators, need
%               to find the courage to pursue polices that acknowledge
%               reality, encourage students to have a good time while
%               drinking responsibly, and don't simply aim to insulate
%               ourselves from lawsuits and helicopter parents.

Before the meeting concluded, however, Quincy House tutor Seth Moulton '01
raised an uncomfortable question highlighting the tensions inherent to
Harvard's approach to campus drinking. He pointed out that a decade ago
Harvard provided undergraduates opportunities to drink---both legally and
illegally---at supervised, house-sponsored events like Stein Club where
underage drinking used to be tolerated in the hopes that students might learn
how to drink responsibly.

And at least, in my case, the approach worked: I did learn how to drink. I
remember distinctly the night I encountered a blockmate in the basement of
Lowell House. ``There's alcohol in the Grille,'' he said. ``It's very
strange.'' He was underage. I was underage. There was alcohol available in
the grille. But it wasn't strange. It was the comfortable public alcohol that
brings friends together, not the furtive thing taken to shot-pounding
extremes behind closed doors. Eventually, hoisting a pint under the watchful
eyes of house tutors became a very normal experience for me, and today I am a
better drinker because of the (illegal) lessons I learned at Stein Club.

Today Harvard doesn't tolerate those grey areas---BAT teams patrol Stein
Club. We are good at focusing on student safety as long as it doesn't incur
any University liability. When it does, we turn away from those we are trying
to educate and claim to care about, pursuing confrontational and ineffective
strategies. Three examples follow to illustrate my point: the legal drinking
age, the clubs, and the Harvard-Yale Tailgate.

If Harvard cared about student safety it would lead or join campaigns to
return the drinking age to 18. The mixture of legal and underage drinkers on
campus is the source of most of our difficulties with alcohol. Students don't
want tutors patrolling their parties in fear that some or most of the
attendees are underage. And tutors don't drop by parties or Stein Club
because confronting underage drinking maroons us between the reality of
campus drinking and our own liabilities. Lowering the drinking age would also
allow all students to patronize the bars within walking distance, where
trained---and liable---bartenders monitor their clientele as they meter out
drinks.

Why won't Harvard support lowering the legal drinking age? The line we hear
repeatly is that 21 is Massachusetts state law, and that is true. But ``Don't
Ask, Don't Tell'' was Federal law and Harvard made it's disagreement with
that misguided policy extremely clear and extremely public. Trying to defend
the current drinking age is hopeless, but Harvard won't publicly
decriminalize underage drinking on campus. It won't publicly eliminate
underage drinking as an institutional disciplinary priority. It won't promise
to protect house masters, tutors, and other residential life staff in case of
lawsuits emanating from alcohol-related accidents on campus, freeing us to
focus on student safety. Harvard won't even join the 135 other institutions
as a signatory to the Amethyst Initiative, which only goes as far as
encouraging ``informed and unimpeded debate on the 21 year-old drinking
age.'' In the face of this silence, it's hard to come to any other conclusion
that Harvard won't act because it cares about liability more than safety.
Because Harvard worries that a lowered drinking age will mean more drinking,
more accidents, and more lawsuits.

If Harvard cared about student safety it would aggressively patrol the Final
Clubs surrounding campus. After all, this is where a large amount of
unmonitored and underage student drinking takes place. Why won't it? Because
the clubs aren't Harvard property, and improving campus safely might annoy
club alumni and their deep pockets.

If Harvard cared about student safety it would increase the length of the
Harvard-Yale tailgate. Every year the tailgate gets less safe, and every year
the University response is to reduce the length or restrict the kinds of
alcohol served. Harvard's persistence in pursuing a failed strategy would be
more impressive if it were less difficult to posit a connection between a
longer tailgate and student safety. The shorter the tailgate, the faster the
drinking; the faster the drinking, the less safe the event. Why won't Harvard
sponsor a more relaxed, more normal Harvard-Yale tailgate? Perhaps because a
shorter tailgate also limits the amount of time that Harvard is sponsoring an
alcohol-driven event, reducing their liability.

At the end of the day it comes down to another key virtue that Seth
mentioned: courage. Courage to reconcile safety v. liability concerns in
favor of safety and students, and take a few unpopular countercultural
stances along the way. A courage the College had ten years ago, and seems to
have lost.

% <wc:end>
